---
title: "tutorial_01_observed_size_spectra"
output: html_document
date: "2024-01-22"
---


```{r}
rm(list =ls()) # clear memory
library(tidyverse)

install.packages(c("tidyverse", "plotly", "remotes", "usethis",
                   "rmarkdown", "rstudioapi"))
#remotes::install_version("mizer", version = "2.3.1")

#remotes::install_github("sizespectrum/mizerExperimental", ref = "tuneMR")
#remotes::install_github("sizespectrum/mizerMR", ref = "364972c")
```

```{r}
#usethis::use_git_config(user.name = "katybbaker", 
                      #  user.email = "katherine.baker@utas.edu.au")


```

#tutorial 1 - observed size spectra
```{r}
download.file("https://github.com/sizespectrum/mizerCourse/raw/main/understand/size-data.rds",
              destfile = "size-data.rds")
```

```{r}
data <- readRDS("size-data.rds")
str(data) #length in centimetres (l) was converted to weight in grams (w) using standard allometric relationship: w = al^b. a and b are species specific parameters 
```
```{r}
unique(data$species)
```
```{r}
p <- ggplot(data) +
    geom_histogram(aes(weight), fill = "blue", colour = "black") +
    scale_x_continuous(name = "Weight [g]") +
    scale_y_continuous(name = "Number of fish")

p
```
#log y axis

```{r}
p + scale_y_log10()
```
#log sized bins

```{r}
(log_breaks <- seq(from = 0, to = 11, by = 1))

(binbreaks <- 2 ^ log_breaks)
```

```{r}
p2 <- ggplot(data) +
    geom_histogram(aes(weight), fill = "blue", colour = "black",
                   breaks = binbreaks) +
    labs(x = "Weight [g]",
         y = "Number of fish") +
    scale_y_log10()
p2
```
#log w axis
```{r}
p2 + scale_x_log10()

```
#now find density by binning by hand

```{r}
data_with_bins <- data |>
    mutate(bin = cut(weight, breaks = binbreaks, right = FALSE,
                     labels = FALSE))
head(data_with_bins)
```


```{r}
#we then group the data by bin and calculate the number of fish in each bin
binned_numbers <- data_with_bins |> 
    group_by(bin) |> 
    summarise(Numbers = n())
binned_numbers
```

#Number density

```{r}
bin_width <- diff(binbreaks)

binned_numbers <- binned_numbers |> 
    mutate(Number_dens = Numbers / bin_width[bin])

```

```{r}
#Let’s make a plot of the number density against weight. Note that we only estimated the number density of weights within each bin, so we only have an average value which we use to determine the height of the curve at the midpoint of the bin. So the plot interpolates between these discrete points (at the midpoint of each bin weights) by straight lines to produce a continuous curve.

bin_midpoints <- 2 ^ (log_breaks[-length(log_breaks)] + 1/2)

binned_numbers$bin_midpoint = bin_midpoints

ggplot(binned_numbers) +
    geom_line(aes(x = bin_midpoint, y = Number_dens)) +
    scale_x_continuous(name = "Weight [g]") +
    scale_y_continuous(name = "Number density")
```
```{r}
ggplot(binned_numbers) +
    geom_line(aes(x = bin_midpoint, y = Number_dens)) + 
    scale_x_log10(name = "Weight [g]") + 
    scale_y_log10(name = "Number density")
```
#fitting a power law

```{r}
(model <- lm(log(Number_dens) ~ log(bin_midpoint), data = binned_numbers))
```
```{r}
ggplot(binned_numbers, aes(x = bin_midpoint, y = Number_dens)) +
    geom_line() + 
    scale_x_log10(name = "Weight [g]") + 
    scale_y_log10(name = "Number density") +
    geom_smooth(method = 'lm')
```
```{r}
w_min = min(data$weight)
lambda <- 1 + nrow(data) / sum(log(data$weight / w_min))
lambda
```
#Biomass density

```{r}
#Above we first calculated the number of fish in each weight bin and then divided by the width of each bin to obtain the average number density in each bin. Exactly analogous to that, we can calculate the biomass of all the fish in each weight bin and divide that by the width of each bin to obtain the average biomass density in each bin. Note, that number density and biomass density will give quite different results, because there are lots of fish in small bins but their total biomass may not be very large. So in the code below we will now sum weight and not numbers

binned_biomass <- data_with_bins |> 
    group_by(bin) |> 
    summarise(Biomass = sum(weight)) |>
    mutate(Biomass_dens = Biomass / bin_width[bin]) |>
    mutate(bin_midpoint = bin_midpoints)

```
```{r}
ggplot(binned_biomass, aes(x = bin_midpoint, y = Biomass_dens)) +
    geom_line() + 
    scale_x_log10(name = "Weight [g]") + 
    scale_y_log10(name = "Biomass density") +
    geom_smooth(method = 'lm')
```
```{r}
(model <- lm(log(Biomass_dens) ~ log(bin_midpoint), data = binned_biomass))
```
#Densities in log weight

```{r}
log10_breaks <- log10(binbreaks)
log10_bin_width <- diff(log10_breaks)
log10_bin_width
```

```{r}
# remember how we calculated densities by dividing numbers and biomasses 
# by bin_widths
binned <- data_with_bins |> 
    group_by(bin) |> 
    summarise(Numbers = n(),
              Biomass = sum(weight)) |>
    mutate(bin_midpoint = bin_midpoints,
           Number_dens = Numbers / bin_width[bin],
           Biomass_dens = Biomass / bin_width[bin])

# and now we add densities in log weight by dividing by log_bin_width
binned <- binned |> 
    mutate(Number_dens_log_w = Numbers / log10_bin_width[bin],
           Biomass_dens_log_w = Biomass / log10_bin_width[bin])

ggplot(binned) +
    geom_line(aes(x = bin_midpoint, y = Number_dens),
              colour = "black") +
    geom_line(aes(x = bin_midpoint, y = Biomass_dens),
              colour = "blue") +
    geom_line(aes(x = bin_midpoint, y = Number_dens_log_w),
              colour = "black", linetype = "dashed") +
    geom_line(aes(x = bin_midpoint, y = Biomass_dens_log_w),
              colour = "blue", linetype = "dashed") +
    scale_x_log10(name = "Weight [g]") +
    scale_y_log10(name = "Density")
```
Summary about the slopes:
The slope of the number density is approximately -2, the slope of the biomass density and the number density in log weight is approximately -1, and the slope of the biomass density in log weight is approximately 0, i.e., the biomass density in log weight is approximately constant.


#size spectra of individual species

```{r}
p <- ggplot(data) +
    geom_density(aes(weight, stat(count), colour = species), adjust = 4) +
    geom_density(aes(weight, stat(count)), colour = "black", lwd = 1.2, adjust = 4) +
    scale_x_continuous(trans = "log10", name = "Weight [g]") +
    scale_y_continuous(trans = "log10", limits = c(1, NA), name = "Number density in log w")

plotly::ggplotly(p)
```
3 takeaways from plot (if it worked):

1. Different species have very different size spectra.
2. The estimates of the species size spectra are not very reliable because we do not have very good data.
3. The community size spectrum looks more regular than the species size spectra.

We will discuss species size spectra more in the next tutorial, where we will look at them with the help of the mizer model.

#Summary and recap:

1) It is very useful to know how many organisms of different sizes there are. This is what size spectra show.

2) We can represent size spectra in different ways. One is to bin the data and plot histograms. The drawback is that the height of the bars in a histogram depend on our choice of bins. A bin from 1 to 2g will have fewer individuals than a bin from 1 to 10g.

3) To avoid the dependence on bin sizes, we use densities, where the total number of individuals or the total biomass of individuals in each bin are divided by the width of the bin. We refer to these as the number density or the biomass density respectively.

3) The number density looks very different from the biomass density. There will be a lot of very small individuals, so the number density at small sizes will be large, but each individual weighs little, so their total biomass will not be large.

5) When we work with a logarithmic weight axis, then it is natural to use densities in log weight, where the numbers of individuals in each bin are divided by the width of the bin on the log axis. We refer to these as the number density in log weight or the biomass density in log weight respectively.

6) Community size spectra are approximately power laws. When displayed on log-log axes, they look like straight lines. The slope of the number density is approximately -2, the slope of the biomass density and the number density in log weight is approximately -1, and the slope of the biomass density in log weight is approximately 0, i.e., the biomass density in log weight is approximately constant. The latter is called the Sheldon spectrum.

7) The individual species size spectra are not approximately power laws and thus do not look like straight lines on log-log axes. The approximate power law only emerges when we add all the species spectra together.



